Pure Continuation Marks

### 2013-04-11 paper.tex
For TFP.

\title{Instant Continuation Marks--Just Add $\lambda$}

@@

We present direct and CPS-like transformations from the call-by-value $\lambda$-calculus augmented with continuation marks to the pure call-by-value $\lambda$-calculus. We discuss how such transformations simplify the construction of compilers which treat continuation marks correctly. We document how Redex, a domain-specific language for exploring language semantics, aided the discovery of correct transformations. We verify the transformations with the proof of a meaning-preservation theorem. Finally, we apply the transformation to JavaScript.
\end{abstract}

@@

We begin by further explaining continuation marks in section \ref{sec-continuation-marks}. We then formalize them via the introduction of a core and enhanced language in section \ref{sec-core-and-extension}. We discuss the necessary properties of a meaningful transformation between these languages. We present two transformations of continuation marks--one direct and one CPS-like--to the call-by-value \lc\ in section \ref{sec-transformation}. This is followed by randomized testing using a lightweight mechanized model in section \ref{sec-testing} and a proof of a meaning-preservation theorem in section \ref{sec-proof}. We finally apply it to JavaScript in section \ref{sec-javascript}.

@@

The evaluation of the factorial of 2, expressed by \scheme|(fac 2)|, proceeds as
\begin{schemedisplay}
(fac 2)
(if (zero? 2) ... (wcm 2 (* 2 (fac (- 2 1)))))
(wcm 2 (* 2 (fac (- 2 1))))
(wcm 2 (* 2 (fac 1)))
(wcm 2 (* 2 (if (zero? 1) ... (wcm 1 (* 1 (fac (- 1 1)))))))
(wcm 2 (* 2 (wcm 1 (* 1 (fac (- 1 1))))))
(wcm 2 (* 2 (wcm 1 (* 1 (fac 0)))))
(wcm 2 (* 2 (wcm 1 (* 1 (if (zero? 0) (begin (print (ccm)) 1) ...)))))
(wcm 2 (* 2 (wcm 1 (* 1 (begin (print (ccm)) 1)))))
\end{schemedisplay}
At this point, the term \scheme|(ccm)| is evaluated. This happens, in effect, by walking the continuation and building a list of the current continuation marks which appear in the \scheme'mark-value' position of a \scheme'wcm' term. In this example, \scheme|(ccm)| evaluates to \scheme|'(1 2)|. Evaluation continues as
\begin{schemedisplay}
(wcm 2 (* 2 (wcm 1 (* 1 (begin (print '(1 2)) 1)))))
(wcm 2 (* 2 (wcm 1 (* 1 (begin 1)))))
(wcm 2 (* 2 (wcm 1 (* 1 1))))
(wcm 2 (* 2 (wcm 1 1)))
(wcm 2 (* 2 1))
(wcm 2 2)
2
\end{schemedisplay}
The visible effect of the call is
\begin{verbatim}
> (fac 2)
(1 2)
2
\end{verbatim}

@@

The evaluation of the factorial of 2, here expressed by \scheme|(fac 2 1)|, proceeds as
\begin{schemedisplay}
(fac 2 1)
(if (zero? 2) ... (wcm 2 (fac (- 2 1) (* 2 1))))
(wcm 2 (fac (- 2 1) (* 2 1)))
(wcm 2 (fac 1 (* 2 1)))
(wcm 2 (fac 1 2))
(wcm 2 (if (zero? 1) ... (wcm 1 (fac (- 1 1) (* 1 2)))))
(wcm 2 (wcm 1 (fac (- 1 1) (* 1 2))))
\end{schemedisplay}
At this point, there is one \scheme'wcm' directive directly nested within another. Behaviorally, the outer one is collapsed \emph{before} evaluation of the inner body proceeds. In essence, it is overwritten. Following this, evaluation continues as
\begin{schemedisplay}
(wcm 1 (fac (- 1 1) (* 1 2)))
(wcm 1 (fac 0 (* 1 2)))
(wcm 1 (fac 0 2))
(wcm 1 (if (zero? 0) (begin (print (ccm)) 2) ...))
(wcm 1 (begin (print (ccm)) 2))
(wcm 1 (begin (print '(1)) 2))
(wcm 1 (begin 2))
(wcm 1 2)
2
\end{schemedisplay}
The visible effect of the call is
\begin{verbatim}
> (fac 2 1)
(1)
2
\end{verbatim}
from which the overwriting which occurred is evidenced the singleton list obtained by \scheme|(ccm)|.

@@

We have not added lists as primitive values to $\cm$ but instead encode lists with other terms in the language. By not relying on a particular characterization of lists in the target language, the dependencies of the transformation remain few, and thus widen its possible application. However, this makes lists themselves subject to the transformation which complicates their manifestation in the core language. The effect of this is that the transformation must deal with the list of continuation marks at the transformed level.

@@

The formal transformation of application follows the \scheme'let' version exactly except the definitions of \scheme'rator-value' and \scheme'rand-value' are folded directly in.
\begin{schemedisplay}
(lambda (flags)
  (lambda (marks)
    (((((C[rator-expr] FALSE) marks)
       ((C[rand-expr] FALSE) marks))
      flags)
     marks)))
\end{schemedisplay}
\end{schemedefn}

@@

\subsection{Definition of $\mathcal{C}$ in CPS}

Some compilers transform source to continuation-passing style (CPS) in the compilation process. The use of CPS as an intermediate language enables a host of optimization analyses and makes the implementation of otherwise-heavyweight features trivial, such as first-class continuations \cite{appel2007compiling}.

We present a continuation mark transformation integrated in CPS.

\begin{schemedefn}{\scheme|Ccps[(rator-expr rand-expr)]|}
\begin{schemedisplay}
(lambda (kont)
   (lambda (flag)
     (lambda (marks)
       (((Ccps[rator-expr]
          (lambda (rator-value)
            (((Ccps[rand-expr]
               (lambda (rand-value)
                 ((((rator-value rand-value) kont) flag) marks)))
              FALSE)
             marks)))
         FALSE)
        marks))))
\end{schemedisplay}
\end{schemedefn}

\begin{schemedefn}{\scheme|C[(wcm mark-expr body-expr)]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (((Ccps[mark-expr]
          (lambda (mark-value) 
            ((lambda (rest-marks) 
               (((Ccps[body-expr] kont) TRUE) Chcps[((CONS mark-value) rest-marks)]))
             ((flag Chcps[(SND marks)]) marks))))
        FALSE)
       marks))))
\end{schemedisplay}
\end{schemedefn}

\begin{schemedefn}{\scheme|Ccps[(ccm)]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (kont marks))))
\end{schemedisplay}
\end{schemedefn}

\noindent
\begin{minipage}{0.5\linewidth}
\begin{schemedefn}{\scheme|Ccps[v]|=\scheme|Ccps[(lambda (x) e)]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (kont (lambda (x) Ccps[e])))))
\end{schemedisplay}
\end{schemedefn}
\end{minipage}
\begin{minipage}{0.5\linewidth}
\begin{schemedefn}{\scheme|Ccps[x]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (kont x))))
\end{schemedisplay}
\end{schemedefn}
\end{minipage}
\\
\\

We include corresponding definitions for $\mathcal{C}'$ and $\hat{\mathcal{C}}$.
\\
\\
\noindent
\begin{minipage}{0.5\linewidth}
\begin{schemedefn}{\scheme|Cpcps[(lambda (x) e)]|}
\begin{schemedisplay}
(lambda (x) Ccps[e])
\end{schemedisplay}
\end{schemedefn}
\end{minipage}
\begin{minipage}{0.5\linewidth}
\begin{schemedefn}{\scheme|Chcps[p]|}
\begin{schemedisplay}
(((Ccps[p] (lambda (x) x)) FALSE) Cpcps[nil])
\end{schemedisplay}
\end{schemedefn}
\end{minipage}

@@

\scheme'redex-check' generates random terms according to the grammar of the given language (\scheme'lambdacm') and category (\scheme'e') in search of counterexamples to the predicate. It gradually increases the size of the terms it generates, which we found useful in obtaining minimal test cases. We subjected both the direct and CPS transformation to random testing and each eventually withstood 10,000 random tests. Interestingly, no incorrect transformation withstood more than 500 random tests before failing.

@@

With a reasonable intuition and definitions hardened by random testing, we prove correctness of the continuation-passing variant of the transform, $\mathcal{C}_{\mathrm{cps}}$.

Because of space limitations, we offer only a sketch of the proof here. The full proof can be found in a technical report\footnote{\texttt{http://students.cs.byu.edu/\textasciitilde{}krgerman/esop2013-technical-report.pdf}}.

@@

\begin{theorem}
For all programs $p\in\lambda_{cm}$, $\Chcps{\mathrm{eval}_{cm}(p)}=\mathrm{eval}_{v}(\Chcps{p})$.
\end{theorem}

We prove this by overloading $\mathcal{C}_{\mathrm{cps}}$ to accommodate first evaluation contexts and then context-term pairs (denoted in unorthodox style) by
\begin{schemedefn}{\scheme|Ccps[E[e]]|}
\begin{schemedisplay}
(((Ccps[e] Ccps[E]) xiE) Cpcps[chiE])
\end{schemedisplay}
\end{schemedefn}
\noindent
which allows us to formally relate $E[e]$ and $\Ccps{E[e]}$. We then prove a lemma
\begin{lemma}[Simulation]
For all contexts $E\in\lambda_{cm}$ and expressions $e\in\lambda_{cm}$, $E[e]\cmrr E'[e']\implies\C{E[e]}\lvrrs\C{E'[e']}$
\end{lemma}
by induction over contexts $E$ and terms $e$. This lemma justifies the evolution of the evaluation context and each rule in the reduction relation except rule \ref{reduction}. We preserve this rule with
\begin{lemma}[Substitution]
\label{lem:substitution}
For all $e,x,v\in\lambda_{cm}$, $\C{e[x\leftarrow v]}=\C{e}[x\leftarrow \Cp{v}]$.
\end{lemma}

@@

Our macro-style approach would be more useful if deployed as a proper macro, but this requires a hygienic macro system which many languages lack. \texttt{sweet.js} is just such a system for JavaScript in early stages and is ideal for our transformation.

Although we used some level of mechanization in the proof of the theorem, a formal proof assistant, such as Coq \cite{barras1997coq}, would increase confidence out of the gate that the transformation was correct.

>>>>

\title{Pure Continuation Marks}

@@

We present a transformation from the call-by-value $\lambda$-calculus augmented with continuation marks to the pure call-by-value $\lambda$-calculus. We discuss how such transformations simplify the construction of compilers which treat continuation marks correctly. We document how Redex, a domain-specific language for exploring language semantics, aided the discovery of correct transformations. We offer the sketch proof of a meaning-preservation theorem. Finally, we apply the transformation to JavaScript.
\end{abstract}

@@

We begin by further explaining continuation marks in section \ref{sec-continuation-marks}. We then formalize them via the introduction of a core and enhanced language in section \ref{sec-core-and-extension}. We discuss the necessary properties of a meaningful transformation between these languages. We present a transformation of continuation marks to the call-by-value \lc\ in section \ref{sec-transformation}. This is followed by randomized testing using a lightweight mechanized model in section \ref{sec-testing} and a sketch proof of a meaning-preservation theorem in section \ref{sec-proof}. We finally apply it to JavaScript in section \ref{sec-javascript}.

@@

The effect of evaluating the factorial of 2, expressed as \scheme|(fac 2)|, is 
\begin{verbatim}
> (fac 2)
(1 2)
2
\end{verbatim}

@@

The effect of evaluating the factorial of 2, here expressed by \scheme|(fac 2 1)|, is
\begin{verbatim}
> (fac 2 1)
(1)
2
\end{verbatim}
from which the overwriting which occurred is evidenced the singleton list obtained by \scheme|(ccm)|.

@@

We have not added lists as primitive values to \cm but instead encode lists with other terms in the language. By not relying on a particular characterization of lists in the target language, the dependencies of the transformation remain few, and thus widen its possible application. However, this makes lists themselves subject to the transformation which complicates their manifestation in the core language. The effect of this is that the transformation must deal with the list of continuation marks at the transformed level.

@@

The formal transformation of application follows the \scheme'let' version exactly except the definitions of \scheme'rator-value' and \scheme'rand-value' are folded directly in.
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    (((((C[rator-expr] FALSE) marks)
       ((C[rand-expr] FALSE) marks))
      flag)
     marks)))
\end{schemedisplay}
\end{schemedefn}

@@

@@

\scheme'redex-check' generates random terms according to the grammar of the given language (\scheme'lambdacm') and term category (\scheme'e') in search of counterexamples to the predicate. It gradually increases the size of the terms it generates, which we found useful in obtaining minimal test cases. We subjected the transformation to random testing. We used the specific counterexamples to guide modifications to the transform until it  withstood 10,000 random tests. Interestingly, no incorrect transformation withstood more than 500 random tests before failing.

@@

With a reasonable intuition and definitions hardened by random testing, we lightly stetch a proof of the correctness of $\mathcal{C}$.

%Because of space limitations, we offer only a sketch of the proof here. The full proof can be found in a technical report\footnote{\texttt{http://students.cs.byu.edu/\textasciitilde{}krgerman/esop2013-technical-report.pdf}}.

@@

\begin{theorem}
For all programs $p\in\lambda_{cm}$, $\Ch{\mathrm{eval}_{cm}(p)}=\mathrm{eval}_{v}(\Ch{p})$.
\end{theorem}

We prove this by overloading $\mathcal{C}$ to accommodate first evaluation contexts and then context-term pairs (where $E[e]$ represents $(E,e)$) by
\begin{schemedefn}{\scheme|C[E[e]]|}
\begin{schemedisplay}
C[E][((C[e] xiE) Cp[chiE])]
\end{schemedisplay}
\end{schemedefn}
\noindent
which allows us to formally relate $E[e]$ and $\C{E[e]}$.
%For example, $\C{\app{\abs{x}{x}}{\hole}[42]}=\C{\app{\abs{x}{x}}{\hole}}[\app{\app{\C{42}}{\xi(\app{\abs{x}{x}}{\hole})}}{\Cp{\chi(\app{\abs{x}{x}}{\hole})}}]$.

We then prove a lemma
\begin{lemma}[Simulation]
For all contexts $E\in\lambda_{cm}$ and expressions $e\in\lambda_{cm}$, $E[e]\cmrr E'[e']\implies\C{E[e]}\lvrrs\C{E'[e']}$
\end{lemma}
by induction over contexts $E$ and terms $e$. This lemma justifies the evolution of the evaluation context and each rule in the reduction relation except rule \ref{reduction}. We preserve this rule with
\begin{lemma}[Substitution]
\label{lem:substitution}
For all $e,x,v\in\lambda_{cm}$, $\C{e[x\leftarrow v]}=\C{e}[x\leftarrow \Cp{v}]$.
\end{lemma}

@@

%Our macro-style approach would be more useful if deployed as a proper macro, but this requires a hygienic macro system which many languages lack. \texttt{sweet.js} is just such a system for JavaScript in early stages and is ideal for our transformation.

%Although we used some level of mechanization in the proof of the theorem, a formal proof assistant, such as Coq \cite{barras1997coq}, would increase confidence out of the gate that the transformation was correct.
